Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-09-15 main.tex
Edits in statistics section.

At the initiation of each perturbation we log the instananeous steer and roll
angles to characterize the configuration of the bicycle when perturbed. The
gain setting on the balance assist controller indicates if the assistance is
off \(g=0\) or on at two different levels: low \(g=8\) or high \(g=10\). A
recovery from the perturbation is successful if the person neither places their
foot down onto the treadmill surface or their wheel of the bicycle exits the
width of the treadmill belt. We record this as a binary variable \(f\) for
"fall". All measured variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

\subsection{Statistics}
%
We test the hypothesis that the balance assistance controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary fall outcome variable \(f\) that is dependent on several possible
explanatory independent variables, one of which is the binary balance
assistance (on or off). See Table~\ref{tab:stat-model-variables} for all of the
variables.
%
\begin{table}
  \centering
  \caption{Statistical model variables.}
  \label{tab:stat-model-variables}
  \begin{tabular}{lll}
    \toprule
    Variable & Units & Description \\
    \midrule
    \(L\) & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(c\) & integer & order number of perturbation \\
    \(g\) & \si{\newton\second\squared} & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (1.7 or 2.8) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
    \bottomrule
  \end{tabular}
\end{table}

We evaluate this hypothesis using a multivariate logistic regression model that
takes the general form
%
\begin{align}
  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij}) \textrm{.}
\end{align}
%
Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
participant \(i\) which follows a Bernoulli distribution given the probability
\(p_{ij}\) that a fall occurs. The log-odds of the probability is then a linear
function of our independent variables with \(\beta\) as the intercept and
\(\alpha_k\) as the linear coefficients to the independent variables
\(x_{ij}^k\), i.e. all variables in Table~\ref{tab:stat-model-variables} except
\(f\).
%
\begin{align}
  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) =
  \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
  \label{eq:log-regress}
\end{align}

Before fitting the model, we scale each \(x_{ij}^k\) such they they have a mean
of zero and a standard deviation of one by cluster-mean centering, as
recommended by \cite{Enders2007}, with clusters being an individual subject.
The clusters are all data from an individual subject because we are only
interested in the association between the state of the balance-assist system
and the outcome of the perturbation. We expected there to be a variation
between participants in how well they are able to resist the perturbation.
However, this was not true. Cluster-mean centering shows there to be no
variation between participants. This allowed us to stick with a simple
single-level logistic regression model, instead of a multilevel model. This
left us with angular impulse, perturbation order, balance assist state, and
roll \& steer angles at the time of perturbation as independent variables. We
also include interaction effects between the balance assist state and the other
four variables.

We divide the analysis into two separate model evaluations, one for the
6~\si{\kph} \(g=10\) trials and one for the
10~\si{\kph} \(g=8\) trials and we evaluate our hypothesis for
each set of data.

\section{Results}
%
The coefficient estimates for a single-level logistic regression at
6~\unit{\kph} with gain $g=10$ are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors. Larger
angular impulse increases the probability to fall and enduring more
perturbations or having the balance assist on, decrease the probability to
fall.  The associated multiplicative change in odds are shown in
Table~\ref{tab:freq-coefs-6}.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at
    6~\unit{\kilo\meter\per\hour} and gain $g=10$. Multiplicative change in
    odds of predictor variables at 6 \unit{\kilo\meter\per\hour} and gain
    $g=10$ based on frequentist single-level logistic regression.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\)-value & $e^{\alpha_k}$ & 2.5\% & 97.5\% \\
    \midrule
    Intercept & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse & 1.69 & 0.27 & 0.00 & 5.40 & 3.18 & 9.16 \\
    Perturbation order & -0.77 & 0.22 & 0.00 & 0.46 & 0.30 & 0.72 \\
    Balance assist state & -0.64 & 0.27 & 0.02 & 0.53 & 0.31 & 0.89 \\
    Roll angle & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance-assist on $\times$ roll angle & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance-assist on $\times$ steer angle & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance-assist on $\times$ angular impulse & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance-assist on $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at
    10~\unit{\kilo\meter\per\hour} and gain $g=8$.  Multiplicative change in
    odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on
    frequentist single-level logistic regression. and 5\% confidence intervals
  of MCiO.}
  \label{tab:freq-coefs-10}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\)-value & $e^{\alpha_k}$ & 2.5\% & 97.5\% \\
    \midrule
    Intercept                                     & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse                               & 2.39  & 0.29 & 0.00 & 10.92 & 6.23 & 19.13 \\
    Perturbation order                            & -1.16 & 0.21 & 0.00 & 0.31 & 0.21 & 0.48 \\
    Balance-assist on                             & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle                                    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle                                   & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance-assist on $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance-assist on $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance-assist on $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance-assist on $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

At the initiation of each perturbation we log the instananeous steer and roll
angles to characterize the configuration of the bicycle when perturbed. The
gain setting on the balance assist controller indicates if the assistance is
off \(g=0\) or on at two different levels: low \(g=8\) or high \(g=10\). A
recovery from the perturbation is successful if the person neither places their
foot down onto the treadmill surface or their wheel of the bicycle exits the
width of the treadmill belt. We record this as a binary variable \(f\) for
"fall". All measured variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance. The
    remaining measures are time varying.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

\subsection{Statistics}
%
We test our hypothesis that the balance assistance controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary fall outcome variable \(f\) that is dependent on several possible
explanatory independent variables, one of which is the binary balance
assistance (on or off). See Table~\ref{tab:stat-model-variables} for the
statistical model variables.
%
\begin{table}
  \centering
  \caption{Independent and dependent variables used in the statistical model.}
  \label{tab:stat-model-variables}
  \begin{tabular}{llll}
    \toprule
    Variable & Causality & Units & Description \\
    \midrule
    \(L\) & ind. & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(\delta_0\) & ind. & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & ind. & \si{\degree} & roll angle at start of perturbation \\
    \(f\) & dep. & boolean & outcome: did not fall 0 or did fall 1 \\
    \(c\) & ind. & boolean & balance assist: off 0 or on 1 \\
    \(j\) & ind. & integer & order number of perturbation \\
    \bottomrule
  \end{tabular}
\end{table}

We evaluate this hypothesis using a multivariate logistic regression model that
takes the general form
%
\begin{align}
  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij}) \textrm{.}
\end{align}
%
Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
participant \(i\) which follows a Bernoulli distribution given the probability
\(p_{ij}\) that a fall occurs. The log-odds of the probability is then a linear
function of our independent variables with \(\beta\) being the intercept and
\(\alpha_k\) the linear coefficients to the \(K\) independent variables
\(x_{ij}^k\), i.e. all variables in Table~\ref{tab:stat-model-variables} except
\(f\).
%
\begin{align}
  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) =
  \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
  \label{eq:log-regress}
\end{align}

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \cite{Enders2007}, with clusters being an
individual subject. The clusters are chosend as all data from an individual
subject because we are only interested in the association between the state of
the balance-assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation. However, this was not true. Cluster-mean centering
shows there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
set of data.

\section{Results}
%
The coefficient estimates for a single-level logistic regression at
6~\unit{\kph} with gain $g=10$ are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors. Larger
angular impulse increases the probability to fall and enduring more
perturbations or having the balance assist on, decrease the probability to
fall.  The associated multiplicative change in odds are shown in
Table~\ref{tab:freq-coefs-6}.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at
    6~\unit{\kilo\meter\per\hour} and gain $g=10$. Multiplicative change in
    odds of predictor variables at 6 \unit{\kilo\meter\per\hour} and gain
    $g=10$ based on frequentist single-level logistic regression.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\)-value & $e^{\alpha_k}$ & 2.5\% & 97.5\% \\
    \midrule
    Intercept & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\) & 1.69 & 0.27 & 0.00 & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)  & -0.77 & 0.22 & 0.00 & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(c\) & -0.64 & 0.27 & 0.02 & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)& -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\) & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance-assist on $\times$ roll angle & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance-assist on $\times$ steer angle & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance-assist on $\times$ angular impulse & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance-assist on $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates at 10~\kph and gain $g=8$.
    Multiplicative change in odds of predictor variables at 10~\kph based on
    frequentist single-level logistic regression. and 5\% confidence intervals
    of MCiO.}
  \label{tab:freq-coefs-10}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\)-value & $e^{\alpha_k}$ & 2.5\% & 97.5\% \\
    \midrule
    Intercept                                     & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse \(L\)                         & 2.39  & 0.29 & 0.00 & 10.92 & 6.23 & 19.13 \\
    Perturbation order \(j\)                      & -1.16 & 0.21 & 0.00 & 0.31 & 0.21 & 0.48 \\
    Balance-assist on \(c\)                       & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle \(\phi_0\)                         & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle \(\delta_0\)                      & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance-assist on $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance-assist on $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance-assist on $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance-assist on $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}
